### 2024-05-28 manuscript/manuscript_blinded_revised.qmd
Intermediate commit of revised manuscript files

The predominant view in developmental psychology is that exposure to adversity impairs cognitive abilities.
This view is supported by decades of research showing that people living in high-adversity contexts tend to score lower on a variety of cognitive tests [@hackman_2010; @ursache_2016].
Recent adaptation-based perspectives, however, have argued that people from adversity may also develop intact, or enhanced, abilities for solving problems in high-adversity contexts [@ellis_2017; @frankenhuis_2013].
Adaptation- and deficit-based perspectives are considered complimentary.
For instance, adversity may impair some cognitive processes, yet enhance others.
Despite their compatibility, few studies have investigated how the interplay of impaired and enhanced abilities shapes performance.
In this paper, across three preregistered online experiments, we used cognitive modeling to derive a process-level understanding of the association between childhood adversity and performance on the Flanker Task, a popular measure of cognitive control [@ridderinkhof_2021].

@@

It is well-established that early-life adversity is associated with deficits in the ability to inhibit distracting, goal-irrelevant information [@hackman_2010; @ursache_2016].
One of the leading paradigms in this literature is the Flanker Task [@eriksen_1974].
On this task, participants typically see five arrows in a horizontal orientation, and are asked to indicate the direction of the central arrow.
The flanking arrows point in the opposite direction on half of the trials, leading to interference that participants must inhibit.
Slower performance on the Flanker has been documented for children and adults with lower childhood socioeconomic status [SES\; @farah_2006; @mezzacappa_2004; @noble_2005] and more environmental unpredictability [@fields_2021; @mittal_2015].
These findings are typically interpreted as indicating a deficit in the ability to inhibit distractions.

Nonetheless, some recent studies suggest that growing up in adversity may also be associated with improved abilities such as attention shifting [@fields_2021; @mittal_2015; @young_2022; but see @nweze_2021].
Some studies found deficit patterns on the Flanker alongside enhancements on other aspects of attention within the same participants.
For example, in one study, children with lower SES performed less well on the Flanker, but outperformed children from higher SES on orienting their attention to salient cues and, peripheral cues [@mezzacappa_2004].
Similarly, children with more caregiver switches (an indicator of unpredictability) were found to perform less well on the Flanker, but outperformed children with fewer caregiver switches on shifting their attention between different task goals [@fields_2021].

Performance on attention tasks could reflect developmental adaptation to adverse environments [@blair_2012; @dangiulli_2012; @frankenhuis_2020; @mittal_2015].
In unpredictable or threatening conditions, the ability to detect salient peripheral information (e.g., distant noises or approaching individuals) could help to more quickly detect and act on potential threats.
Over time, cognitive adaptations to such conditions could result in a general tendency to use a more diffuse scope of attention, leading to an enhanced ability to keep track of the broader environment.
In line with this hypothesis, people with lower SES respond more strongly to auditory distractors [@giuliano_2018; @stevens_2009; @hao_2022] and are faster to orient their attention to peripheral visual information [@mezzacappa_2004].
While potentially adaptive, a more diffuse scope of attention could come at the cost of lowered ability to ignore irrelevant distractors.
This could compromise more long-term goal-directed behavior, especially in chaotic environments (e.g., a noisy classroom or a busy street).

Thus, lowered performance on tasks like the Flanker could reflect either a cognitive impairment or a difference in attentional strategies.
Distinguishing between these two possibilities is challenging for two reasons.
First, few studies in the adversity literature have measured performance differences on different attention tasks within the same individual [@mezzacappa_2004; @mittal_2015].
Thus, it is unclear whether lowered inhibition is related to enhanced processing of peripheral information in people from from adverse backgrounds.
Second, performance on inhibition tasks is---beyond the ability to inhibit distractors---also influenced by other factors, such as a person's general processing speed and response caution [@hedge_2022; @loffler_2022].
This means that lowered performance on inhibition tasks does not necessarily reflect inhibition difficulties.
In other words, we should consider cognitive processes other than ability when drawing inferences based on inhibition tasks.

@@

An important issue, therefore, is that several processes are involved in performance on the Flanker, and standard assessments using raw performance measures (response times, accuracy rates) mostly fail to distinguish between them.
For example, performance differences on the Flanker could indicate that someone experiences more (or less) flanker interference, generally processes less (or more) efficiently, or responds with less (or more) caution.
To understand how adversity affects performance, we need to be able to separate the difference processes that make up performance.

Formal cognitive models such as the Drift Diffusion Model [DDM\; @forstmann_2016; @ratcliff_1998; @ratcliff_2008; @ratcliff_2015; @wagenmakers_2009] provide a potential solution.
The DDM estimates cognitive processes at different stages of the decision-making process (See Figure 1a).
It represents decision-making on binary decision-making tasks as a process in which people accumulate information until one response is sufficiently favored over the other.
These two response options are represented as opposing boundaries.
One boundary corresponds to the correct response and the other to the incorrect response.
When information accumulation reaches one of the two boundaries, the corresponding response is executed.

The DDM translates trial-level response times (RTs) and accuracy into three distinct cognitive processes.
The speed of information accumulation is captured in a parameter called the *drift rate*.
Higher drift rates are associated with faster responses and higher mean accuracy.
Response caution is modeled through the *boundary separation*; that is, the width between the two boundaries.
Larger boundary separation is associated with larger RTs and higher accuracy (i.e., sacrificing speed to increase accuracy).
Finally, *non-decision time* represents the time it takes to prepare for the task at the start of the trial (before information processing) and the time it takes to execute a response (after a response boundary has been reached).
Longer non-decision times are associated with larger RTs, without influencing accuracy.

The Shrinking Spotlight (SSP) model is an extension of the standard DDM to account for attention processes on the Flanker [@grange_2016; @white_2018a; @white_2018b; @white_2011].
The SSP model assumes that attention resembles a spotlight that is normally distributed over the Flanker arrows (with a particular starting *attentional width*).
Over time, people narrow their attention down to the central arrow (at a rate defined by the *shrinking rate*), thereby gradually decreasing interference from irrelevant information (See Figure 1b).
Prior work has defined the amount of distractor *interference* by dividing the attentional width by the shrinking rate [@white_2018a].
People may experience less interference either by starting with a narrower attentional width, and/or by more rapidly shrinking their attention down to the target arrow.
Finally, performance is also influenced by the *perceptual input* strength; that is, how well someone can process the arrows in general.
Note that typical interpretations of lowered raw Flanker performance are in terms of the amount of interference that someone experiences, and not in terms of the strength of perceptual input.
Here, we are interested in how childhood adversity is associated with both interference and the strength of perceptual input.

@@

```{r}
#| label: figure1
#| fig.width: 7
#| fig.height: 9
#| dpi: 600
#| fig-cap: | 
#|   **Figure 1.** A visual overview of the Drift Diffusion Model (DDM) and the Shrinking Spotlight Model (SSP). Panel A: The DDM assumes that people go through three distinct stages when they perform cognitive tasks with two forced response options. In a first preparation phase, they visually encode the relevant stimuli. In a second decision phase, people accumulate information in favor of one decision over the other (e.g., pressing left vs. right) until the decision boundary for either the correct or incorrect response is reached. Each jagged line represents this information accumulation process on a single trial. In a third execution phase, people execute the motor response.  The model estimates four parameters that reflect distinct cognitive processes (printed in italic): (1) The drift rate represents the average rate of evidence accumulation towards the correct decision boundary and measures processing speed; (2) The non-decision time represents both the time spent encoding the stimuli and executing the response; (3) The boundary separation represents how far apart has “set” their decision boundaries, and is a measure of the person’s level of response caution; (4) The starting point represents the starting point of the decision process and measures the level of response bias (not considered here). Panel B: The SSP is an extension of the standard DDM including additional parameters to capture attentional processes involved in Flanker performance. Each stimulus arrow provides a certain strength of perceptual input (*p*). On incongruent trials, the perceptual input of flanking arrows is coded negatively (-*p*). Attention is assumed to be normally distributed over the arrows with a certain attentional width. Over time, attention is narrowed down toward the central arrow at a rate determined by the shrinking rate, thereby gradually lowering the interference caused by the flanking arrows. The drift rate in the SSP model is the sum of the perceptual input of each arrow multiplied by the attention allotted to each arrow. As attention for the flanking arrows decreases over time, the drift rate is assumed to increase over time (contrary to the standard DDM, which assumes a linear drift rate).

@@

The overarching goal of our studies is to understand the attentional and processing styles that people develop in conditions of adversity.
We focus on measures of exposure to violence and environmental unpredictability.
Previous research shows that these two types of adversity are on the one hand associated with improved attention shifting and working memory updating [@fields_2021; @mittal_2015; @young_2018; @young_2022], and on the other hand with lowered inhibition and working memory capacity [@fields_2021; @mittal_2015; @young_2018].
We conducted three online studies: one pilot and two follow-up studies.
Using cognitive modeling, we unpack Flanker performance in comparison to other tasks that require externally focused attention (Pilot), across visual processing manipulations (Study 1), and in terms of tendencies for holistic versus local processing (Study 2).

We used an incremental preregistration approach across studies (for all preregistrations, code and materials, see <https://anonymous.4open.science/r/CogDev_AnonLink>).
For each study, we preregistered primary (i.e., hypothesis-driven) and exploratory analyses.
The main text addresses the primary analyses involving violence exposure and the exploratory analyses involving environmental unpredictability.
We describe the other exploratory analyses in the supplemental materials (section 2).
For an overview of all deviations from the preregistrations, see section 4 of the supplemental materials.

# Pilot

In the Pilot, our goal was to understand how childhood adversity relates to performance on tasks with different attentional demands.
Participants completed self-report measures of childhood adversity and three cognitive tasks.
These tasks measured inhibition, attention for peripheral cues, and attention for subtle changes.
We expected violence exposure to be positively associated with a present-oriented attention style.
Specifically, we expected participants with more violence exposure to be better at detecting peripheral stimuli and subtle changes.
We expected this would result in a higher drift rate (faster speed of information accumulation) or shorter non-decision times (faster attention orientation, among other things), but not necessarily with differences in boundary separation (response caution).
Second, we expected that participants with more violence exposure would be worse at ignoring distracting peripheral stimuli.
We expected this would result in more experienced interference (as derived from the SSP model).

@@

Participants were 565 people from the United States aged between 18 and 30 recruited on Prolific Academic ([https://www.prolific.co](https://www.prolific.co)) (See Table 1 for demographic data).
The sample was balanced on sex.
We used the MacArthur's ladder, included in Prolific's prescreening battery, for assessing perceived SES to ensure about half of the sample came from lower-SES backgrounds (which we defined as a score of 4 or below).
Participants were eligible if they spoke fluent English and did not report color-blindness.

@@

Prior to analyzing the data, we applied our preregistered exclusion criteria.
First, we excluded participants who did not complete the full study; second, those who had incomplete data on any of the attention tasks; third, those who missed both attention check items; fourth, those who had suspicious response patterns (e.g., consistently endorsing high response options even though some items were reverse coded).
Fifth, on a trial-level, we excluded any trials with reaction times \< 250 ms or \> 3500 ms [@ratcliff_2015].
Participants with more than 10 trials removed were completely excluded from the analyses.\
The final sample consisted of `r nrow(pilot_data)` participants.

@@

**Flanker Task.** The Flanker measures selective attention and response inhibition [@eriksen_1974].
The Flanker began with eight practice trials, followed by 64 test trials.
On each trial, participants saw a set of five arrows pointing either left or right.
Participants were instructed to indicate the direction of the central arrow while ignoring the flanking arrows to the left and right.
All trials included black arrows against a white background.
On the *congruent* trials (50%), the flanking arrows pointed in the same direction as the central arrow.
On the *incongruent* trials (50%), the arrows pointed in the opposite direction.
The arrows were randomly presented in the top-half or bottom-half of the screen.
Each trial started with a fixation cross (1000 ms), after which the arrows were visible until a response was given.
Participants received performance feedback during the practice trials, but not during the test block.

**Cued Attention Task.** The Cued Attention Task was an adapted version of the Posner task, which measures the speed of attention for peripheral cues [@posner_1980].
The Cued Attention Task began with eight practice trials, followed by 64 test trials.
On each trial, a left- or right-pointing arrow was presented in one of eight random locations at 300 pixels from the center of the screen.
Participants were instructed to indicate the direction of the arrow by pressing either the left- or right arrow key on their keyboard.
All trials included a black cue and arrow against a white background.
On *cued* trials (50%), a cue ('\*') preceded the arrow in the exact same location.
On *neutral* trials (50%), the cue preceded the arrow, but appeared at the center of the screen (not where the arrow would appear).
Thus, the cue was perfectly predictive of the target location on cued trials, but provided no predictive information about the location of the arrow on neutral trials.
Each trial started with a fixation cross at the center of the screen for 1000 ms. Then, the cue appeared for 250 ms, followed by the target arrow, until a response was given.

**Change Detection Task.** The Change Detection Task measures the ability to detect subtle spatial changes.
The Change Detection Task started with five practice trials followed by 50 test trials.
On each trial, participants saw five colored circles (red, light-blue, dark-blue, yellow, and purple) against a gray background, each with a radius of 15 pixels.
Each circle was located in a semi-random location around the central fixation cross.
The location of each circle was sampled within a pre-specified area of 50-by-50 pixels to prevent overlap.
Participants had 1000 ms to memorize the locations of the five circles.
Then, the circles disappeared for 500 ms and then reappeared.
On *change* trials (50%), one of the circles had moved to another location with a fixed displacement of 40 pixels in a 360 degree direction.
On *no change* trials (50%), all circles were still in the same location.
Participants were instructed to indicate whether all circles were still in the same location or one of the circles had changed location.
The displacement of *one* circle was the only potential difference on each trial;

**DDM/SSP parameters.** We analyzed the Flanker with the SSP model [@grange_2016; @white_2018a; @white_2018b; @white_2011], using the *flankr* package [@grange_2016].
For each participant, the SSP provided us with estimates of: (1) strength of perceptual input (general quality of information that participants get from the arrows), (2) interference (initial attention width divided by the speed at which attention is narrowed down to the central arrow), (3) non-decision time (combination of speed of initial stimulus encoding and response execution), and (4) boundary separation (response caution).
Our focus on interference as a ratio between attention width and shrinking rate was a deviation from the preregistration, as we initially planned to investigate both aspects of attention separately.
However, we discovered that both parameters in isolation were unreliable because of an inherent trade-off, while the ratio did provide a stable measure.
This was supported in a simulation study by @white_2018a showing that the ratio measure is reliable.
See the supplemental materials (section 3) for a comparison between the preregistered and the updated analyses.

For the Change Detection Task and the Cued Attention Task, we used a hierarchical Bayesian implementation of the standard DDM (HDDM).
For each participant, the HDDM provided us with estimates of: (1) drift rate (speed of information accumulation; analogous to strength of perceptual input in the SSP model, except that drift rate is time-invariant), (2) non-decision time (same as SSP model), and (3) boundary separation (same as SSP model).
The hierarchical Bayesian fitting procedure was a deviation from the preregistration, in which we planned to use Maximum Likelihood (ML) estimation.
There were several issues with estimating DDM parameters for the Cued Attention task, which we later discovered were caused specifically by ML.
An important difference between HDDM and ML is that HDDM uses the group information to inform individual parameter estimates, whereas ML models are fitted to each individual separately.
This generally improves the accuracy of the estimation.
See the supplemental materials (section 3) for an overview of the fit procedure and model fit across all models.

@@

Finally, we included four objective measures of unpredictability before age 13: 1) "How often did you move?"; 2) "How many adults lived in your home on average?"; 3) "How many romantic partners did your mother have (not counting your father)?"; 4) "How many romantic partners did your father have (not counting your mother)?".

@@

### Primary analyses

For the Cued Attention Task and Flanker RTs, we used linear mixed effects models to test violence exposure x task condition (sum-coded) interactions on mean RTs and each DDM parameter.
All mixed effects models included a random intercept for participants.
For the Change Detection Task and Flanker SSP parameters, we used linear regression models to test the main effect of adversity on mean RTs and each DDM/SSP parameter.
To meet model assumptions of normally distributed residuals, reaction times were log-transformed.
Analyses involving interference (Flanker) and boundary separation (all tasks) parameters violated the assumption of normally distributed residuals.
For boundary separation, we solved this using log-transformation.
For interference, non-normality was caused by extreme outliers (\>3.2SD), which we excluded from the analyses.

@@

Table 3 summarizes the results.
More violence exposure was associated with lower strength of perceptual input under `r pilot_main_eff_ddm_list$Flanker_p$p_sum_Main`% of multiverse specifications (although the median 95% CI interval contained zero).
We additionally found a significant main effect of violence exposure on interference under `r pilot_main_eff_ddm_list$Flanker_interference$p_sum_Main`% of multiverse specifications, such that more violence exposure was associated with less interference.
This was contrary to our expectation that people exposed to adversity would have more difficulties dealing with interference from irrelevant distractors.

>>>>

The predominant view in developmental psychology is that exposure to adversity---defined as prolonged exposure to intense stress---impairs cognitive abilities.
This view is supported by decades of research showing that people living in high-adversity contexts tend to score lower on a variety of cognitive tests [@hackman_2010; @ursache_2016].
Recent adaptation-based perspectives, however, have argued that people from adversity may also develop intact, or enhanced, abilities for solving problems in high-adversity contexts [@ellis_2017; @frankenhuis_2013].
Adaptation- and deficit-based perspectives are considered complementary.
For instance, adversity may impair some cognitive processes, yet enhance others.
Despite their compatibility, few studies have investigated how the interplay of impaired and enhanced abilities shapes performance.
Across three preregistered online experiments, we used cognitive modeling to derive a process-level understanding of the association between childhood adversity and performance in the Flanker task, a popular measure of cognitive control [@ridderinkhof_2021].

@@

It is well-established that early-life adversity is associated with deficits in the ability to inhibit distracting, goal-irrelevant information [@hackman_2010; @ursache_2016].
One of the leading paradigms in this literature is the Flanker task [@eriksen_1974].
On this task, participants typically see five arrows in a horizontal orientation, and are asked to indicate the direction of the central arrow.
The flanking arrows point in the opposite direction on half of the trials, leading to interference that participants must inhibit.
Slower performance in the Flanker task has been documented for children and adults with more environmental unpredictability [@fields_2021; @mittal_2015].
These findings are typically interpreted as indicating a deficit in the ability to inhibit distractions.

Similar associations have been documented for factors that increase the risk of adversity exposure, such as lower socioeconomic status [SES\; @farah_2006; @mezzacappa_2004; @noble_2005].
Although people living in low-SES conditions experience more adversity, on average, we do not regard low SES itself as a form of adversity. 
First, SES and adversity can affect cognitive abilities through different mechanisms (e.g., education versus physiological stress). 
Second, people with low SES have diverse experiences, both positive and negative, even if adversity is more common in this group.

Some recent studies suggest that growing up in adversity may also be associated with improved abilities such as attention shifting [@fields_2021; @mittal_2015; @young_2022; but see @mezzacappa_2004; @nweze_2021].
Some studies found deficit patterns on inhibition tasks alongside enhancements on other aspects of attention within the same participants.
For example, one study found that young adults with more childhood unpredictability committed more errors on an Antisaccade task (a measure of inhibition), but more efficiently switched their attention between tasks on an attention shifting task [@mittal_2015].
Similarly, children with more caregiver switches (an indicator of unpredictability) experienced more interference in the Flanker task (based on RTs), but outperformed children with fewer caregiver switches on shifting their attention between different task goals [@fields_2021].

Performance on attention tasks could reflect developmental adaptation to adverse environments [@blair_2012; @dangiulli_2012; @frankenhuis_2020; @mittal_2015].
In unpredictable or threatening conditions, the ability to detect salient peripheral information (e.g., distant noises or approaching individuals) could help to more quickly detect and act on potential threats.
Over time, cognitive adaptations to such conditions could result in a general tendency to use a more diffuse scope of attention, leading to an enhanced ability to keep track of the broader environment.
In line with this hypothesis, people with lower SES respond more strongly to auditory distractors [@giuliano_2018; @stevens_2009; @hao_2022] and are faster to orient their attention to peripheral visual information [@mezzacappa_2004].
While potentially adaptive, a more diffuse scope of attention could come at the cost of lowered ability to ignore irrelevant distractors.
This could compromise longer-term goal-directed behavior, especially in chaotic environments (e.g., a noisy classroom or a busy street).

Thus, lowered performance on tasks like the Flanker task could reflect either a cognitive impairment or a difference in attentional strategies.
Distinguishing between these two possibilities is challenging for two reasons.
First, few studies in the adversity literature have measured performance differences on different attention tasks within the same individual [@mezzacappa_2004; @mittal_2015].
Thus, it is unclear whether lowered inhibition is related to enhanced processing of peripheral information in people from from adverse backgrounds.
Second, performance on inhibition tasks is---beyond the ability to inhibit distractors---also influenced by other factors, such as a person's general processing speed and response caution [@hedge_2022; @loffler_2024].
This means that lowered performance on inhibition tasks does not necessarily reflect inhibition difficulties.
In other words, we should consider cognitive processes other than ability when drawing inferences based on inhibition tasks.

@@

An important issue, therefore, is that several processes are involved in performance in the Flanker task, and standard assessments using raw performance measures (response times, accuracy rates) mostly fail to distinguish between them.
For example, performance differences in the Flanker task could indicate that someone experiences more (or less) distractor interference, generally processes less (or more) efficiently, or responds with less (or more) caution.
To understand how adversity affects performance, we need to be able to separate the difference processes that make up performance.

Formal cognitive models such as the Drift Diffusion Model [DDM\; @forstmann_2016; @ratcliff_1998; @ratcliff_2008; @ratcliff_2015; @wagenmakers_2009] provide a potential solution.
The DDM estimates explicitly models the cognitive processes underlying the decision-making (See Figure 1a).
It represents decision-making on binary decision-making tasks as a process in which people accumulate information until one response is sufficiently favored over the other.
These two response options are represented as opposing boundaries.
One boundary corresponds to the correct response and the other to the incorrect response (note that in some research designs, the boundaries may be coded as the two choice options instead, for example, when the question is whether people classify a certain class of stimuli (e.g., angry faces) more efficiently than another class of stimuli (e.g., happy faces)).
When the accumulated information reaches one of the two boundaries, the corresponding response is executed.

The DDM translates trial-level response times (RTs) and accuracy into three distinct cognitive processes.
The speed of information accumulation is captured in a parameter called the *drift rate*.
Higher drift rates are associated with faster responses and higher mean accuracy.
Response caution is modeled through the *boundary separation*; that is, the width between the two boundaries.
Larger boundary separation is associated with larger RTs and higher accuracy (i.e., sacrificing speed to increase accuracy).
*Non-decision time* represents the time it takes to prepare for the task at the start of the trial (before information accumulation starts) and the time it takes to execute a response (after a response boundary has been reached).
Longer non-decision times are associated with larger RTs, without influencing accuracy.
Finally, the *starting point* represents a potential bias towards one of two responses, with a biased decision-making process starting closer to one boundary relative to the other boundary.

The Shrinking Spotlight (SSP) model is an extension of the standard DDM to account for attention processes in the Flanker task [@grange_2016; @white_2018a; @white_2018b; @white_2011].
The SSP model assumes that attention resembles a spotlight that is normally distributed over the Flanker task arrows (with a particular starting *attentional width*).
Over time, people narrow their attention down to the central arrow (at a rate defined by the *shrinking rate*), thereby gradually decreasing interference from irrelevant information [cf\. @eriksen_1986\; see Figure 1b).
Prior work has defined the amount of distractor *interference* by dividing the attentional width by the shrinking rate [@white_2018a].
People may experience less interference either by starting with a narrower attentional width, and/or by more rapidly shrinking their attention down to the target arrow.
Finally, performance is also influenced by the *perceptual input* strength; that is, how well someone can process the arrows in general.
Note that typical interpretations of lowered raw Flanker task performance are in terms of the amount of interference that someone experiences, and not in terms of the strength of perceptual input.

The DDM and SSP model share many assumptions. 
Both provide identical estimates of boundary separation and non-decision time. 
The main difference is the decision-making phase. 
The DDM assumes that the quality of the information is the same across the entire trial. 
In contrast, the SSP model assumes that the quality of the information improves across the trial, as attention becomes gradually focused more on the central arrow. 
The simpler assumption of the DDM makes the model broadly applicable, but less precise for conflict tasks. 
The SSP model is more precise, but applicable only to the Flanker task. 
Previous studies have successfully applied the DDM to Flanker task data [e.g., @loffler_2024; @vermeent_2024a].
However, unlike the DDM, the SSP model affords testing hypotheses about the association between adversity and attentional interference in the Flanker task.
Specifically, we are interested in how childhood adversity is associated with both interference and the strength of perceptual input.
Finally, the SSP model is one of several completing diffusion models developed to explain performance on conflict tasks [@hubner_2010; @ulrich_2015].
We focus exclusively on the SSP model because it performs well with relatively few trials per participant relative to other conflict diffusion models [@white_2018a].

@@

```{r}
#| label: figure1
#| fig.width: 7
#| fig.height: 9
#| dpi: 600
#| fig-cap: | 
#|   **Figure 1.** A visual overview of the Drift Diffusion Model (DDM) and the Shrinking Spotlight Model (SSP). Panel A: The DDM assumes that people go through three distinct stages when they perform cognitive tasks with two forced response options. In a first preparation phase, they visually encode the relevant stimuli. In a second decision phase, people accumulate information in favor of one decision over the other (e.g., pressing left vs. right) until the decision boundary for either the correct or incorrect response is reached. Each jagged line represents this information accumulation process on a single trial. In a third execution phase, people execute the motor response.  The model estimates four parameters that reflect distinct cognitive processes (printed in italic): (1) The drift rate represents the average rate of evidence accumulation towards the correct decision boundary and measures processing speed; (2) The non-decision time represents both the time spent encoding the stimuli and executing the response; (3) The boundary separation represents how far apart a person has “set” their decision boundaries, and is a measure of the person’s level of response caution; (4) The starting point (not considered here) represents a potential bias towards one of two responses, with a biased decision-making process starting closer to one boundary relative to the other boundary. Panel B: The SSP is an extension of the standard DDM including additional parameters to capture attentional processes involved in Flanker task performance. Each stimulus arrow provides a certain strength of perceptual input (*p*). On incongruent trials, the perceptual input of flanking arrows is coded negatively (-*p*). Attention is assumed to be normally distributed over the arrows with a certain attentional width. Over time, attention is narrowed down toward the central arrow at a rate determined by the shrinking rate, thereby gradually lowering the interference caused by the flanking arrows. The drift rate in the SSP model is the sum of the perceptual input of each arrow multiplied by the attention allotted to each arrow. As attention for the flanking arrows decreases over time, the drift rate is assumed to increase over time (contrary to the standard DDM, which assumes a linear drift rate).

@@

The overarching goal of our studies is to understand the attentional and processing styles that people develop in conditions of adversity.
We focus on measures of exposure to violence and environmental unpredictability.
Previous research shows that these two types of adversity are on the one hand associated with improved attention shifting and working memory updating [@fields_2021; @mittal_2015; @young_2018; @young_2022], and on the other hand with lowered inhibition and working memory capacity [@fields_2021; @mittal_2015; @young_2018].
We conducted three online studies: one pilot study and two follow-up studies.
Using cognitive modeling, we unpack Flanker task performance in comparison to other tasks that require externally focused attention (Pilot study), across visual processing manipulations (Study 1), and in terms of tendencies for holistic versus local processing (Study 2).

We used an incremental preregistration approach across studies (for all preregistrations, data, code and materials, see <https://anonymous.4open.science/r/CogDev_AnonLink>).
For each study, we preregistered confirmatory (i.e., hypothesis-driven) and exploratory analyses.
The main text addresses the confirmatory analyses involving violence exposure and the exploratory analyses involving environmental unpredictability.
We describe the other exploratory analyses in the supplemental materials (section 2).
For an overview of all deviations from the preregistrations, see section 4 of the supplemental materials.

# Pilot study

In the Pilot study, our goal was to understand how childhood adversity relates to performance on tasks with different attentional demands.
Participants completed self-report measures of childhood adversity and three cognitive tasks (Flanker task, Cued Attention task, and Change Detection task).
These tasks measured inhibition, attention for peripheral cues, and attention for subtle changes.
In line with the idea that exposure to adversity may lead to a more diffuse scope of attention, we expected people with more violence exposure to be better at detecting peripheral stimuli and subtle changes.
We expected this would result in a higher drift rate (faster speed of information accumulation) or shorter non-decision times (faster attention orientation, among other things), but not necessarily with differences in boundary separation (response caution).
In contrast, we expected that participants with more violence exposure would be worse at ignoring distracting peripheral stimuli.
We expected this would result in more experienced interference (as derived from the SSP model).

@@

Participants were 565 people from the United States aged between 18 and 30 recruited on Prolific Academic ([https://www.prolific.co](https://www.prolific.co)) (See Table 1 for demographic data).
The sample was balanced on sex.
We used the MacArthur's ladder, included in Prolific's prescreening battery, for assessing perceived SES to ensure about half of the sample came from lower-SES backgrounds (which we defined as a score of 4 or below).
Participants were eligible if they spoke fluent English and did not report color-blindness.
We obtained ethical approval from the first author's institute (anonymized for peer review).

@@

Prior to analyzing the data, we applied our preregistered exclusion criteria.
First, we excluded participants who did not complete the full study; second, those who had incomplete data on any of the attention tasks; third, those who missed both attention check items; fourth, those who had suspicious response patterns (e.g., consistently endorsing high response options even though some items were reverse coded).
Fifth, on a trial-level, we excluded any trials with reaction times \< 250 ms or \> 3500 ms [@ratcliff_2015].
Participants with more than 10 trials removed were completely excluded from the analyses.
The final sample consisted of `r nrow(pilot_data)` participants.

@@

**Flanker task.** The Flanker task measures selective attention and response inhibition [@eriksen_1974].
The Flanker task began with eight practice trials, followed by 64 test trials.
On each trial, participants saw a set of five arrows pointing either left or right.
Participants were instructed to indicate the direction of the central arrow by pressing the respective arrow keys, while ignoring the flanking arrows to the left and right.
All trials included black arrows against a white background.
in the *congruent* trials (50%), the flanking arrows pointed in the same direction as the central arrow.
in the *incongruent* trials (50%), the arrows pointed in the opposite direction.
The arrows were randomly presented in the top-half or bottom-half of the screen.
Each trial started with a fixation cross (1000 ms), after which the arrows were visible until a response was given.
Participants received performance feedback during the practice trials, but not during the test block.

**Cued Attention task.** The Cued Attention task was an adapted version of the Posner task, which measures the speed of attention for peripheral cues [@posner_1980].
The Cued Attention task began with eight practice trials, followed by 64 test trials.
On each trial, a left- or right-pointing arrow was presented in one of eight random locations at 300 pixels from the center of the screen.
Participants were instructed to indicate the direction of the arrow by pressing either the left- or right arrow key on their keyboard.
All trials included a black cue and arrow against a white background.
On *cued* trials (50%), a cue ('\*') preceded the arrow in the exact same location.
On *neutral* trials (50%), the cue preceded the arrow, but appeared at the center of the screen (not where the arrow would appear).
Thus, the cue was perfectly predictive of the target location on cued trials, but provided no predictive information about the location of the arrow on neutral trials.
Each trial started with a fixation cross at the center of the screen for 1000 ms. Then, the cue appeared for 250 ms, followed by the target arrow, until a response was given.

**Change Detection task.** The Change Detection task measures the ability to detect subtle spatial changes.
The Change Detection task started with five practice trials followed by 50 test trials.
On each trial, participants saw five colored circles (red, light-blue, dark-blue, yellow, and purple) against a gray background, each with a radius of 15 pixels.
Each circle was located in a semi-random location around the central fixation cross.
The location of each circle was sampled within a pre-specified area of 50-by-50 pixels to prevent overlap.
Participants had 1000 ms to memorize the locations of the five circles.
Then, the circles disappeared for 500 ms and then reappeared.
On *change* trials (50%), one of the circles had moved to another location with a fixed displacement of 40 pixels in a 360 degree direction.
On *no change* trials (50%), all circles were still in the same location.
Participants were instructed to indicate whether all circles were still in the same location or one of the circles had changed location by pressing the left- or right-arrow key.
The displacement of *one* circle was the only potential difference on each trial;

**DDM/SSP parameters.** We analyzed Flanker task performance with the SSP model [@grange_2016; @white_2018a; @white_2018b; @white_2011], using the *flankr* package [@grange_2016].
For each participant, the SSP provided us with estimates of: (1) strength of perceptual input (general quality of information that participants get from the arrows), (2) interference (initial attention width divided by the speed at which attention is narrowed down to the central arrow), (3) non-decision time (combination of speed of initial stimulus encoding and response execution), and (4) boundary separation (response caution).
We always fixed the starting-point to the midpoint between the two boundaries, as modeling bias makes little sense when the boundaries correspond to correct and incorrect responses (as is the case here), rather than the distinct response options.
Our focus on interference as a ratio between attention width and shrinking rate deviated from the preregistration, as we initially planned to investigate both aspects of attention separately.
However, we discovered that both parameters in isolation were unreliable because of an inherent trade-off, while the ratio did provide a stable measure.
This was supported in a simulation study by @white_2018a showing that the ratio measure is reliable.
See the supplemental materials (section 3) for a comparison between the preregistered and the updated analyses.

For the Change Detection task and the Cued Attention task, we used a hierarchical Bayesian implementation of the standard DDM (HDDM).
For each participant, the HDDM provided us with estimates of: (1) drift rate (speed of information accumulation; analogous to strength of perceptual input in the SSP model, except that drift rate is time-invariant), (2) non-decision time (same as in the SSP model), and (3) boundary separation (same as in the SSP model).
The hierarchical Bayesian fitting procedure was a deviation from the preregistration, in which we planned to use Maximum Likelihood (ML) estimation.
There were several issues with estimating DDM parameters for the Cued Attention task, which we later discovered were caused specifically by ML.
An important difference between HDDM and ML is that HDDM uses the group information to inform individual parameter estimates, whereas ML models are fitted to each individual separately.
The hierarchical approach generally improves generally improves the accuracy of the estimation.
See the supplemental materials (section 3) for an overview of the fit procedure and model fit across all studies.

@@

Finally, we included four objective measures of unpredictability before age 13: 1) "How often did you move?"; 2) "How many adults lived in your home on average?"; 3) "How many romantic partners did your mother have (not counting your father)?"; 4) "How many romantic partners did your father have (not counting your mother)?".
Previous studies have found associations between (subsets of) these measures and subjective measures of adversity as well as with developmental outcomes [@belsky_2012; @ellis_2009; @young_2022].

@@

**Confirmatory analyses.** For the Cued Attention and Flanker task RTs, we used linear mixed effects models to test violence exposure x task condition (sum-coded) interactions on mean RTs (calculated separately for each condition) and each DDM parameter.
All mixed effects models included a random intercept for participants.
For the Change Detection and Flanker task SSP parameters, we used linear regression models to test the main effect of adversity on mean RTs and each DDM/SSP parameter.
We did not analyze accuracy rates as these were close to ceiling for the Flanker and Cued Attention task.
To meet model assumptions of normally distributed residuals, mean reaction time were log-transformed, separately for the congruent and incongruent condition.
Analyses involving interference (Flanker task) and boundary separation (all tasks) parameters violated the assumption of normally distributed residuals.
For boundary separation, we solved this using log-transformation.
For interference, non-normality was caused by extreme outliers (\>3.2SD), which we excluded from the analyses.

@@

Table 3 summarizes the results.
In the flanker task, More violence exposure was associated with lower strength of perceptual input under `r pilot_main_eff_ddm_list$Flanker_p$p_sum_Main`% of multiverse specifications (although the median 95% CI interval contained zero).
We additionally found a significant main effect of violence exposure on interference under `r pilot_main_eff_ddm_list$Flanker_interference$p_sum_Main`% of multiverse specifications, such that more violence exposure was associated with less interference.
This was contrary to our expectation that people exposed to adversity would have more difficulties dealing with interference from irrelevant distractors.

Participants with more exposure to childhood violence were slower in the Cued Attention task, which was mainly related to a higher level of response caution (boundary separation).
in the Change Detection task, more childhood violence exposure was associated with slower speed of information processing (drift rate) under `r pilot_main_eff_ddm_list$Change_v$p_sum_Main`% of multiverse specifications, but not with longer RTs.
These results were not in line with our expectation that people from adversity would perform better on cognitive tasks that require a broad, present-focused attention style.
